---
title: "Logistic Regression with Random Intercepts"
author: "Will Townes"
date: "May 7, 2016"
output: html_document
---

```{r}
library(rstan)
library(reshape2)
library(plyr)
library(lme4)
rstan_options(auto_write=TRUE)
options(mc.cores=parallel::detectCores())
set.seed(39)
```

### Model Specification

The outcome for $i^{th}$ observation in $k^{th}$ cluster is binary $Y_{ki}$ with $P(Y_{ki}=1)=g(X_{ki}\beta+\gamma_k)$ where $\beta$ are the fixed effects, $X_{ki}$ is the vector of covariates, and $\gamma_k$ is a random intercept shared among all observations in the cluster. In traditional mixed models, we assume $\gamma_k\sim\mathcal{N}(0,\sigma^2_\gamma)$ but in the Bayesian context we can specify more flexible distributions for $\gamma_k$.

### Generating Simulated Data

```{r}
expit<-function(x){1/(1+exp(-x))}
gen_data<-function(beta_true){
  K<-100 #clusters
  n<-10 #observations/cluster
  sigma<-2 #random effects standard deviation
  dat<-data.frame(id=factor(rep(1:K,each=n)))
  x1<-rbinom(K,1,.5)
  dat$x1<-rep(x1,each=n) #cluster-specific covar
  dat$x2<-rep(seq(from=-4.5,to=4.5),K) #within-cluster covar
  dat<-within(dat,eta<-beta_true[1]+x1*beta_true[2]+x2*beta_true[3])
  #random intercepts
  bdat<-data.frame(id=factor(1:K),b=rnorm(K,0,sigma)) 
  dat<-join(dat,bdat,by="id")
  dat<-within(dat,mu<-expit(eta+b))
  dat$y<-rbinom(K*n,1,dat$mu)
  return(dat[,c("id","x1","x2","y")])
}
dat_obs<-gen_data(c(-2,1,.5)) #specify true fixed effects
dat_obs2<-gen_data(c(5,-1,1.5)) #alternative dataset
```

First we will use the standard frequentist approach (maximum likelihood) to get a baseline for comparison.

```{r}
frq<-glmer(y~x1+x2+(1|id),data=dat_obs,family=binomial)
summary(frq)
```

Recall the true fixed effects are $-2,1,.5$ and the random effect standard deviation is $2$.

### Stan Program

Run the stan program for random intercepts logistic regression. There is no need for data in a list like in the example on their website, since stan searches the calling environment. The stan code is in a separate file. Commonly, one may specify a normal distribution with mean zero and variance $\sigma^2$ as the random effects distribution, and then place an inverse gamma conjugate prior on $\sigma^2$ with shape $a$ and scale $b$. However, this scheme can be analytically marginalized out if $\sigma^2$ is not of intrinsic interest. If $\sigma^2\sim\mathcal{IG}(a,b)$ then set $\nu=2a$ and $\tau^2=a/b$ so that $\sigma^2\sim\mathcal{IG}(\nu/2,\nu \tau^2/2)$. This is equivalent to a scaled inverse chi-squared prior, with the property that $\tau^2=E[\sigma^2]$. Analytically integrating out $\sigma^2$, we obtain a marginal Student's t prior for the random effects with degrees of freedom $\nu$, mean zero, and scale parameter $\tau^2$.

```{r}
dat2stan<-function(dat){
  #convenience for converting dataframe to list of needed parameters
  X<-as.matrix(dat[,c("x1","x2")])
  id<-as.numeric(dat$id)
  list(N=nrow(X),D=ncol(X),K=length(unique(id)),y=dat$y,X=X,id=id)
}
stmod<-stan_model(file="glmm_logi.stan")
system.time(stfit<-sampling(stmod,data=dat2stan(dat_obs),pars=c("alpha","beta","sigma"),iter=1000))
summary(stfit)$summary[1:4,c("mean","2.5%","97.5%")]
```

Now that the model has compiled, let's see if it saves time to run it again on a different dataset.

```{r eval=FALSE}
system.time(stfit<-sampling(stmod,data=dat2stan(dat_obs2),pars=c("alpha","beta","sigma"),iter=1000))
summary(stfit)$summary[1:4,c("mean","2.5%","97.5%")]
```

The amount of time is the same, and it appears to be accurate. Note that the combination of *stan_model* and *sampling* functions is the same thing as the usual *stan()* function, except by separating out computation and sampling we don't have to recompile for every new dataset if the model is the same.

## Stan Front-End Packages

As shown by [Kazuki Yoshida](http://rpubs.com/kaz_yos/glmm1), there are a number of convenient front-ends for stan. Let's try some of those out and see if results are comparable.

### Package brms

```{r}
library(brms)
system.time(stfit2 <- brm(formula=y~x1+x2+(1|id), data=dat_obs, family="bernoulli",iter=1000))
summary(stfit2)
```

Looks pretty good! We can also extract the generated stan model. It looks like this:

```{r}
(brm_mod<-stfit2$model)
```

This is useful if we want to run the same model multiple times. However, it does not allow us to run the model on multiple datasets without re-compiling. Each new dataset must be recompiled even if the stan code is the same. If the below code were run, it would show that when "fit" is specified, all other arguments including data are ignored and the model fits the first dataset and fails to fit the second dataset. In other words, don't do the following!

```{r eval=FALSE}
system.time(stfit2b<-brm(fit=stfit2,data=dat_obs2))
summary(stfit2b)
system.time(stfit2c<-brm(fit=stfit2,data=dat_obs2))
summary(stfit2c)
```

We can use other brm functions though to separate out data transformations, compiling the stan program, etc. to save compilation time

```{r}
brm_mod<-stan_model(model_code=stfit2$mod)
brm_dat<-make_standata(y~x1+x2+(1|id),data=dat_obs,family="bernoulli")
system.time(stfit2b<-sampling(brm_mod,brm_dat,pars=c("b_Intercept","b","sd_1"),iter=1000))
summary(stfit2b)$summary[,c("mean","2.5%","97.5%")]
```

Now we try switching the dataset without switching the compiled model

```{r}
brm_dat<-make_standata(y~x1+x2+(1|id),data=dat_obs2,family="bernoulli")
system.time(stfit2b<-sampling(brm_mod,brm_dat,pars=c("b_Intercept","b","sd_1"),iter=1000))
summary(stfit2b)$summary[,c("mean","2.5%","97.5%")]
```

Excellent, it seems to work without re-compiling and gives reasonable estimates for the parameters.

### Packages rstanarm and MCMCglmm

Another front-end option is shown below.

```{r}
library(rstanarm)
system.time(stfit3<-stan_glmer(formula=y~x1+x2+(1|id), data=dat_obs, family="binomial"))
summary(stfit3)
```

Package brms appears to be faster than rstanarm. Finally, we try package MCMCglmm. I couldn't get it to run.

```{r eval=FALSE}
library(MCMCglmm)
system.time(stfit4<-MCMCglmm(fixed=y~x1+x2, random=~(1:id), family="categorical", data=dat_obs))
summary(stfit4)
```