---
title: "Logistic Regression with Random Intercepts"
author: "Will Townes"
date: "May 7, 2016"
output: html_document
---

```{r}
library(rstan)
library(reshape2)
library(plyr)
library(lme4)
rstan_options(auto_write=TRUE)
options(mc.cores=parallel::detectCores())
set.seed(39)
```

### Model Specification

The outcome for $i^{th}$ observation in $k^{th}$ cluster is binary $Y_{ki}$ with $P(Y_{ki}=1)=g(X_{ki}\beta+\gamma_k)$ where $\beta$ are the fixed effects, $X_{ki}$ is the vector of covariates, and $\gamma_k$ is a random intercept shared among all observations in the cluster. In traditional mixed models, we assume $\gamma_k\sim\mathcal{N}(0,\sigma^2_\gamma)$ but in the Bayesian context we can specify more flexible distributions for $\gamma_k$. Specifically here we consider the model where $\gamma_k$ comes from a two component gaussian mixture model.

### Generating Simulated Data

```{r}
expit<-function(x){1/(1+exp(-x))}
gen_data<-function(beta_true){
  K<-100 #clusters
  n<-10 #observations/cluster
  dat<-data.frame(id=factor(rep(1:K,each=n)))
  x1<-rbinom(K,1,.5)
  dat$x1<-rep(x1,each=n) #cluster-specific covar
  dat$x2<-rep(seq(from=-4.5,to=4.5),K) #within-cluster covar
  dat<-within(dat,eta<-beta_true[1]+x1*beta_true[2]+x2*beta_true[3])
  #random intercepts
  mix_comp<-sample(c(-2,2),K,replace=TRUE)
  bdat<-data.frame(id=factor(1:K),b=rnorm(K,mix_comp,.5)) 
  dat<-join(dat,bdat,by="id")
  dat<-within(dat,mu<-expit(eta+b))
  dat$y<-rbinom(K*n,1,dat$mu)
  return(dat[,c("id","x1","x2","y")])
}
dat_obs<-gen_data(c(-2,1,.5)) #specify true fixed effects
dat_obs2<-gen_data(c(5,-1,1.5)) #alternative dataset
```

First we will use the standard frequentist approach (maximum likelihood) to get a baseline for comparison.

```{r}
frq<-glmer(y~x1+x2+(1|id),data=dat_obs,family=binomial)
summary(frq)
```

Recall the true fixed effects are $-2,1,.5$.

### Stan Program

```{r}
dat2stan<-function(dat){
  #convenience for converting dataframe to list of needed parameters
  X<-as.matrix(dat[,c("x1","x2")])
  id<-as.numeric(dat$id)
  J<-2
  a<-J+2 #close to uniform Dirichlet
  list(N=nrow(X),D=ncol(X),K=length(unique(id)),y=dat$y,X=X,id=id,J=J,a=a)
}
stmod<-stan_model(file="glmm_logi_mix.stan")
system.time(stfit<-sampling(stmod,data=dat2stan(dat_obs),pars=c("fixed_int","beta","theta","mu","sigma"),iter=1000,control=list(adapt_delta=.85,max_treedepth=12)))
summary(stfit)$summary[1:4,c("mean","2.5%","97.5%")]
pairs(stfit,pars="mu") #distinctive bimodal distributions
```

We now try again on a different dataset

```{r}
system.time(stfit<-sampling(stmod,data=dat2stan(dat_obs2),pars=c("fixed_int","beta","theta","mu","sigma"),iter=1000,control=list(adapt_delta=.85,max_treedepth=12)))
summary(stfit)$summary[1:4,c("mean","2.5%","97.5%")]
```